State of the Art in Concurrent Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-20 paper/paper.tex
Paper edits

\begin{abstract}
Priority queues are abstract data structures which store a set of (priority, item) pairs
and allow efficient access to the item with minimal priority. Such queues are an important
element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
path algorithm) and operating system (priority schedulers).

The recent trend towards multiprocessor computing requires new implementations of basic
data structures which are able to be used concurrently and ideally scale well to a large number
of threads. In particular, lock-free structures promise superior scalability by avoiding
the use of blocking synchronization primitives.

@@

Concurrent priority queues have been the subject of research since the 1980s
\cite{ayani1990lr,biswas1987simultaneous,das1996distributed,deo1992parallel,huang1991evaluation,
luchetti1993some,mans1998portable,olariu1991optimal,prasad1995parallel}.
While early efforts have focused mostly on parallelizing Heap structures
\cite{hunt1996efficient}, % TODO: Many more, see \cite{shavit2000skiplist}
more recently priority queues based on \citeauthor{pugh1990skip}'s SkipLists
\cite{pugh1990skip} seem to show more potential \cite{shavit2000skiplist,
sundell2003fast,herlihy2012art,linden2013skiplist}. Current research has also examined
relaxed data structures \cite{wimmer2013data,alistarhspraylist} which trade
the strictness of provided guarantees against improved scalability.

@@

In the following, we examine the evolution of concurrent priority queues suitable for general use,
i.e. unbounded range queues based on widely available atomic primitives such as \ac{CAS}.
Section \ref{sec:concepts} outlines basic concepts and definitions. In Section \ref{sec:hunt},
we cover the \citeauthor{hunt1996efficient} queue as a representative of early heap-based queues using
fine-grained locking to avoid the bottleneck of a single global lock. Lock-free SkipList-based
structures offering better disjoint-access parallelism are discussed in Section \ref{sec:lockfree},
while Section \ref{sec:relaxed} introduces the concept of relaxed data structures and presents two
such priority queues. Finally, related work is presented in Section \ref{sec:related}, and the paper is
concluded in Section \ref{sec:conclusion}.

@@

There have been many publications on more specialized variants of concurrent priority queues
which we did not cover in this paper.

@@

\citeauthor{liu2012lock} present a lock-free, array-based priority queue in \cite{liu2012lock}
using the non-standard \ac{DCAS} and \ac{DCSS} instructions. In addition to the standard
\lstinline|Insert| and \lstinline|DeleteMin| operations, it also provides an \lstinline|ExtractMany|
function which returns the $n$ highest priority items in the queue, and a function that returns
items which ``probably'' have high priority. Other implementations based on non-standard instructions
can be found in \cite{israeli1993efficient,greenwald1999non}.

@@

Another possibility is the restriction of priorities to a particular set, i.e. bounded range priorities.
\citeauthor{shavit1999scalable} present two such data structures in \cite{shavit1999scalable} using
combining funnels and bins of items.

@@

Finally, the principle of relaxation has also been applied to other data structures.
\citeauthor{kirsch2012fast} invented an efficient k-FIFO queue \cite{kirsch2012fast},
% TODO: More cites.

@@

The switch from heaps to SkipLists as the backing data structure highlights how a simple change in
direction can help revitalize an entire field of research. SkipList-based priority queues are the current
state of the art in strict shared-memory concurrent priority queues: they provide strong guarantees
and scale well to up to the ten's of threads in practice. Important limiting factors are contention
at the front of the list and the large number of \ac{CAS} failures.
The \citeauthor{linden2013skiplist} queue is designed
to minimize the latter factor; but the former is inherent to all strict priority queues, which could mean
that the peak performance in such structures has been reached.

>>>>

\begin{abstract}
Priority queues are abstract data structures which store a set of (priority, item) pairs
and allow efficient access to the item with minimal priority. Such queues are an important
element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
path algorithm) and operating system (i.e. priority schedulers).

The recent trend towards multiprocessor computing requires new implementations of basic
data structures which are able to be used concurrently and scale well to a large number
of threads. In particular, lock-free structures promise superior scalability by avoiding
the use of blocking synchronization primitives.

@@

Concurrent priority queues have been the subject of research since the 1980s
\cite{ayani1990lr,biswas1987simultaneous,das1996distributed,deo1992parallel,huang1991evaluation,
luchetti1993some,mans1998portable,olariu1991optimal,prasad1995parallel}.
While early efforts have focused mostly on parallelizing Heap structures
\cite{hunt1996efficient}, % TODO: Many more, see \cite{shavit2000skiplist}
more recently priority queues based on \citeauthor{pugh1990skip}'s SkipLists
\cite{pugh1990skip} seem to show more potential \cite{shavit2000skiplist,
sundell2003fast,herlihy2012art,linden2013skiplist}. Current research has also examined
relaxed data structures \cite{wimmer2013data,alistarhspraylist} which trade
strictness of provided guarantees against improved scalability.

@@

In the following, we investigate the evolution of concurrent priority queues suitable for general use,
i.e. unbounded range queues based on widely available atomic primitives such as \ac{CAS}.
Section \ref{sec:concepts} outlines basic concepts and definitions. In Section \ref{sec:hunt},
we cover the \citeauthor{hunt1996efficient} queue as a representative of early heap-based queues using
fine-grained locking to avoid the bottleneck of a single global lock. Lock-free SkipList-based
structures offering better disjoint-access parallelism are discussed in Section \ref{sec:lockfree},
while Section \ref{sec:relaxed} introduces the concept of relaxed data structures and presents two
such priority queues. Finally, related work is presented in Section \ref{sec:related}, and the paper is
concluded in Section \ref{sec:conclusion}.

@@

There have been many publications on more specialized variants of concurrent priority queues
which were not covered in this paper.

@@

\citeauthor{liu2012lock} present a lock-free, array-based priority queue in \cite{liu2012lock}
using the non-standard \ac{DCAS} and \ac{DCSS} instructions. In addition to the
\lstinline|Insert| and \lstinline|DeleteMin| operations, it also provides an \lstinline|ExtractMany|
function which returns the $n$ highest priority items in the queue, and a function that returns
items which ``probably'' have high priority. Other implementations based on non-standard instructions
can be found in \cite{israeli1993efficient,greenwald1999non}.

@@

Another possibility is the restriction of priorities to a particular set, i.e. bounded range priorities.
\citeauthor{shavit1999scalable} present two such data structures in \cite{shavit1999scalable} using
combining funnels and bins of items. Experimental results have shown strong scalability until at least
256 threads.

@@

Finally, the principle of relaxation has also been applied to other data structures.
\citeauthor{kirsch2012fast} invented an efficient k-FIFO queue \cite{kirsch2012fast},
The trend towards relaxed data structures in general is examined further in
\cite{shavit2011data,kirsch2012incorrect}.

@@

The switch from heaps to SkipLists as the backing data structure highlights how a simple change in
direction can help revitalize an entire field of research. SkipList-based priority queues are the current
state of the art in strict shared-memory concurrent priority queues: they provide strong guarantees
and scale well to up to the tens of threads in practice. Important limiting factors are contention
at the front of the list and the large number of \ac{CAS} failures.
The \citeauthor{linden2013skiplist} queue is designed
to minimize the latter; but the former is inherent to all strict priority queues, which could mean
that the peak performance in such structures has been reached.
